dispRity manual

### 2017-08-09 doc/dispRity-manual.Rmd
Updates following Natalie's issues

Disparity must be measured from ordinated matrices.
These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).

@@

<!-- I removed the semicolons above, it is confusing to novice R users -->

Of course you can use your own data as detailed in the [previous chapter](#which-data).
<!-- I changed the header so may have broken this link -->

@@

We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `aov`):

@@

## Testing for an difference among the time bins
disp_aov <- test.dispRity(disparity_data, test = aov, comparisons = "all")
summary(disp_aov)

@@

 * An ordinated matrix (always!)
 * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix. For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4. Note that elements can be present in multiple groups at once.
 * Your favourite disparity metric (here the sum of variances)

@@

The function `time.subsamples` allows users to divide the ordinated matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.

@@

For the time-slicing method details see @GuillermeSTD. 
<!-- NC: Or potentially this paper we are writing for the PalAss? -->
For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:

@@

 *  `acctran` where the data chosen along the branch is always the one of the offspring
 *  `deltran` where the data chosen along the branch is always the one of the descendant
 *  `punctuated` where the data chosen along the branch is randomly chosen between the offspring or the descendant
 *  `gradual` where the data chosen along the branch is either the offspring or the descendant depending on branch length

<!--NC: Bit puzzled by the offspring/descendent terminology - I'd assume these were the same thing. By descendant do you mean ancestor? I'm confused  -->

@@

Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-between-groups).
This function simply takes the list of elements to put in each group (whether they are the actual element names or their idea).

<!-- NC: Their idea? Huh? Whats that mean? -->

@@

<!-- NC: This is confusing to me. If it's a number > 0, it keeps that number of dimensions, and if < 0 it removes the last dimensions by that %? This is very odd. Please explain -->

```{r, eval=TRUE}
## Removing the last 50% of the dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

## Keeping the first 10 dimensions
boot.matrix(BeckLee_mat50, dimensions = 10)
```

Of course, one could directly supply the subsamples generated above (using <!--  NC: which function ? How is this line different from the statement in the line below? -->) to this function.
It can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.

@@

## Creating time slice subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
                               method = "continuous", model = "gradual", 
                               time = c(120, 80, 40, 0),
                               FADLAD = BeckLee_ages)
## Bootstrapping the time slice subsamples
boot.matrix(time_slices, bootstraps = 100)
```

@@

There are many ways of measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @finlay2015morphological. 
<!-- NC: probably should cite Matt Wills' paper for the centroid thing too as he invented it apparently -->
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.

@@

<!-- NC: We discussed this previously, but I don't like the term levels here. It's confusing because it means something specific wrt factors. Can we change it to dimension-level? That's more descriptive and clearer in my opinion. I have changed it below, see what you think -->

@@

<!-- NC: My terminology changes may have broken this bit -->

```{r}
## Testing whether mean is dimension-level 1
if(make.metric(mean, silent = TRUE) != "dimension-level1") {
    message("The metric is not dimension-level 1.")
}
## Testing whether var is dimension-level 1
if(make.metric(var, silent = TRUE) != "dimension-level1") {
    message("The metric is not dimension-level 1.")
}
```

@@

Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > dimension-level 2 > dimension-level 1).

@@

It is also possible to apply some more *complex* tests that have their own output classes (like `stats::aov`).

<!-- NC: again this  needs to be lm, aov is not appropriate. Will let you fix this -->
The results can then be analysed as usual using the associated `summary` S3 method:

```{r}
## Performing and ANOVA applied to the same data
(slice_aov <- test.dispRity(disparity_time_slices, test = aov,
                            comparisons = "all"))

## The output is a regular ANOVA output
class(slice_aov)

## This output can be summarised using summary
summary(slice_aov)
```

>>>>

Disparity can be estimated from pretty much any matrix.
Classically, however, it is measured from ordinated matrices.
These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).

@@

Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).

@@

We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):

@@

## Testing for an difference among the time bins
disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
summary(disp_lm)

@@

 * An matrix with rows as elements and columns as dimensions (always!)
 * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix. For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4. Note that elements can be present in multiple groups at once.
 * Your favourite disparity metric (here the sum of variances)

@@

The function `time.subsamples` allows users to divide the matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.

@@

For the time-slicing method details see @GuillermeSTD. <!-- @@@ Change cite appropriately! -->
<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:

@@

 *  `acctran` where the data chosen along the branch is always the one of the descendant
 *  `deltran` where the data chosen along the branch is always the one of the ancestor
 *  `punctuated` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
 *  `gradual` where the data chosen along the branch is either the descendant or the ancestor depending on branch length

@@

Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-among-groups).
This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).

@@

```{r, eval=TRUE}
## Using the first 50% of the dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

## Using the first 10 dimensions
boot.matrix(BeckLee_mat50, dimensions = 10)
```

Of course, one could directly supply the subsamples generated above (using `time.subsamples` or `custom.subsamples`) to this function.

@@

## Creating time slice subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
                               method = "continuous", model = "gradual", 
                               time = c(120, 80, 40, 0),
                               FADLAD = BeckLee_ages)

## Bootstrapping the time slice subsamples
boot.matrix(time_slices, bootstraps = 100)
```

@@

There are many ways of measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994. 
<!-- NC: probably should cite Matt Wills' paper for the centroid thing too as he invented it apparently; TG: ok fixed. Though one can argue Euclides invented it before him (and probably Gnut-the-enlighted-cavemen before-->
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.

@@

@@

```{r}
## Testing whether mean is dimension-level 1
if(make.metric(mean, silent = TRUE) != "level1") {
    message("The metric is not dimension-level 1.")
}
## Testing whether var is dimension-level 1
if(make.metric(var, silent = TRUE) != "level1") {
    message("The metric is not dimension-level 1.")
}
```

@@

Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).

@@

It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).

The results can then be analysed as usual using the associated `summary` S3 method:

```{r}
## Performing and linear model applied to the same data
(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
                            comparisons = "all"))

## The output is a regular `lm` output
class(slice_lm)

## This output can be summarised using summary
summary(slice_lm)
```
